---
title: "Linear Mixed Effects Model:  Mutation burden correlation with Age and Sickle disease status"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lme4)

my_theme<-theme(text = element_text(family="Helvetica"),
                axis.text = element_text(size = 5),
                axis.title = element_text(size=7),
                legend.text = element_text(size=5),
                legend.title = element_text(size=7),
                strip.text = element_text(size=7),
                legend.spacing = unit(1,"mm"),
                legend.key.size= unit(5,"mm"))
plots_dir="~/R_work/Gene_therapy/plots_NEJM/"
```

## R Markdown

Import the mutation burden/ age data

```{r import data}
GT_sample_metadata<-readRDS("~/R_work/Gene_therapy/Gene_therapy_for_SCD_NEJM/Data/sample_metadata_full_with_sigs.Rds")
EM_sample_metadata<-read.csv("~/R_work/Mitochondrial_analysis/EM_sample_level_metadata.csv")

combined_df<-bind_rows(GT_sample_metadata%>%
            filter(!is.na(SNV_burden_AR)&Time_point==0)%>%
            dplyr::select(new_ID,Age_at_GT,Sample,SNV_burden_AR,SNV_burden_invitro_removed,INDEL_burden_AR)%>%
            mutate(Disease_status="SCD")%>%
            dplyr::rename("ID"=new_ID,"Age"=Age_at_GT,"SNV_burden_adj"=SNV_burden_AR,"INDEL_burden_adj"=INDEL_burden_AR),
          EM_sample_metadata%>%
            dplyr::select(donor_id,age,PDID,sub_adj,indel_adj)%>%
            mutate(Disease_status="Healthy")%>%
            dplyr::rename("Age"=age,"ID"=donor_id,"Sample"=PDID,"SNV_burden_adj"=sub_adj,"INDEL_burden_adj"=indel_adj)%>%
            mutate(SNV_burden_invitro_removed=SNV_burden_adj)
            )
head(combined_df)
```

## View data on plot with linear regression lines for each category

Plot

```{r , echo=FALSE}
cols=RColorBrewer::brewer.pal(3,"Dark2")
base_plot<-combined_df%>%
  ggplot(aes(x=Age,y=SNV_burden_adj,col=Disease_status))+
  geom_jitter(width=0.25,height=0,alpha=0.25,size=0.5)+
  theme_classic()+
  scale_color_manual(values=cols)+
  labs(x="Age",y="SNV burden (adjusted)",col="Disease\nstatus")+
  my_theme

base_plot+geom_smooth(method="lm",size=0.5)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Experiment with linear models

Modeling including age only

```{r}
lmer.age_only <- lmer(SNV_burden_adj ~ Age + (1|ID), data = combined_df)
summary(lmer.age_only)

p.1<-base_plot+geom_abline(intercept = lmer.age_only@beta[1],slope = lmer.age_only@beta[2])+theme(legend.position = "none")
ggsave(filename=paste0(plots_dir,"lmer_ageonly.pdf"),p.1,width = 2,height=2)

```

Model including age and disease (without any interaction)

```{r}
lmer.age_and_disease <- lmer(SNV_burden_adj ~ Age + Disease_status + (1|ID), data = combined_df)
summary(lmer.age_and_disease)

p.2<-base_plot+
  geom_abline(intercept=lmer.age_and_disease@beta[1],slope=lmer.age_and_disease@beta[2],col=cols[1])+
  geom_abline(intercept=lmer.age_and_disease@beta[1]+lmer.age_and_disease@beta[3],slope=lmer.age_and_disease@beta[2],col=cols[2])+theme(legend.position = "none")
ggsave(filename=paste0(plots_dir,"lmer_age_disease_fixed.pdf"),p.2,width = 2,height=2)
```


Now test for significance of individual SCD patients versus normal by testing each SCD individual together with all the healthy individuals. Here the "Disease_status" term actually refers to the single patient.
Getting p-values is more straight-forward with the lme function from nlme package.

```{r}
all_individual_lmer_list<-lapply(paste0("SCD",1:6),function(test_ID){
  lmer.indidividual=nlme::lme(SNV_burden_adj ~ Age + Disease_status, data = combined_df%>%filter(Disease_status=="Healthy"|ID==test_ID),random = ~ 1|ID)
  return(lmer.indidividual)
})
all_lmer_summaries<-lapply(all_individual_lmer_list,summary)
summary_df<-dplyr::bind_rows(lapply(all_lmer_summaries,function(list) list$tTable["Disease_statusSCD",]))%>%
  mutate(CI_lower=Value-1.96*Std.Error,CI_upper=Value+1.96*Std.Error)%>%
  mutate(ID=paste0("SCD",1:6),.before=1)
print(summary_df)

```

Model including age + an age:disease interaction term only

```{r}
lmer.age_and_disease_int <- lmer(SNV_burden_adj ~ Age + Age:Disease_status + (1|ID), data = combined_df)
summary(lmer.age_and_disease_int)
confint(lmer.age_and_disease_int)

m1 <- nlme::lme(SNV_burden_adj~Age + Age:Disease_status,random=~1|ID,data=combined_df)
anova(m1)

p.3<-base_plot+
  geom_abline(intercept=lmer.age_and_disease_int@beta[1],slope=lmer.age_and_disease_int@beta[2],col=cols[1])+
  geom_abline(intercept=lmer.age_and_disease_int@beta[1],slope=lmer.age_and_disease_int@beta[2]+lmer.age_and_disease_int@beta[3],col=cols[2])
ggsave(filename=paste0(plots_dir,"lmer_age_disease_int.pdf"),p.3,width = 3,height=2)
```

Model including age + disease + an age:disease interaction term

```{r}
lmer.age_and_disease_plus_int <- lmer(SNV_burden_adj ~ Age*Disease_status + (1|ID), data = combined_df)
summary(lmer.age_and_disease_plus_int)
```

##DO THE SAME WITH THE 'IN VITRO REMOVED' VALUES

## View data on plot with linear regression lines for each category

Plot

```{r , echo=FALSE}
combined_df%>%
  ggplot(aes(x=Age,y=SNV_burden_invitro_removed,col=Disease_status))+
  geom_jitter(width=0.25,height=0,alpha=0.25,size=0.5)+
  theme_classic()+
  geom_smooth(method="lm",size=0.5)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Experiment with linear models

Modeling including age only

```{r}
lmer.age_only.ivr <- lmer(SNV_burden_invitro_removed ~ Age + (1|ID), data = combined_df)
summary(lmer.age_only.ivr)

```

Model including age and disease (without any interaction)

```{r}
lmer.age_and_disease.ivr <- lmer(SNV_burden_invitro_removed ~ Age + Disease_status + (1|ID), data = combined_df)
summary(lmer.age_and_disease.ivr)

```

Now test for significance of individual SCD patients versus normal by testing each SCD individual together with all the healthy individuals. Here the "Disease_status" term actually refers to the single patient.
Getting p-values is more straight-forward with the lme function from nlme package.

```{r}
all_individual_lmer_list.ivr<-lapply(paste0("SCD",1:6),function(test_ID){
  lmer.indidividual=nlme::lme(SNV_burden_invitro_removed ~ Age + Disease_status, data = combined_df%>%filter(Disease_status=="Healthy"|ID==test_ID),random = ~ 1|ID)
  return(lmer.indidividual)
})
all_lmer_summaries.ivr<-lapply(all_individual_lmer_list.ivr,summary)
summary_df.ivr<-dplyr::bind_rows(lapply(all_lmer_summaries.ivr,function(list) list$tTable["Disease_statusSCD",]))%>%
  mutate(CI_lower=Value-1.96*Std.Error,CI_upper=Value+1.96*Std.Error)%>%
  mutate(ID=paste0("SCD",1:6),.before=1)
print(summary_df.ivr)

```

Model including age + an age:disease interaction term only

```{r}
lmer.age_and_disease_int.ivr <- lmer(SNV_burden_invitro_removed ~ Age + Age:Disease_status + (1|ID), data = combined_df)
summary(lmer.age_and_disease_int.ivr)

```

Model including age + disease + an age:disease interaction term

```{r}
lmer.age_and_disease_plus_int.ivr <- lmer(SNV_burden_invitro_removed ~ Age*Disease_status + (1|ID), data = combined_df)
summary(lmer.age_and_disease_plus_int.ivr)
```

##NOW DO THE SAME FOR INDELS

```{r , echo=FALSE}
combined_df%>%
  ggplot(aes(x=Age,y=INDEL_burden_adj,col=Disease_status))+
  geom_jitter(width=0.25,height=0,alpha=0.25,size=0.5)+
  theme_classic()+
  geom_smooth(method="lm",size=0.5)
```

Modeling including age only

```{r}
lmer.age_only.indel <- lmer(INDEL_burden_adj ~ Age + (1|ID), data = combined_df)
summary(lmer.age_only.indel)

```

Model including age and disease (without any interaction)

```{r}
lmer.age_and_disease.indel <- lmer(INDEL_burden_adj ~ Age + Disease_status + (1|ID), data = combined_df)
summary(lmer.age_and_disease.indel)

```
Model including age and disease (without any interaction), but applied to each individual in term to see if they have significant differences individually

```{r}
all_individual_lmer_list.indel<-lapply(paste0("SCD",1:6),function(test_ID){
  lmer.indidividual=nlme::lme(INDEL_burden_adj ~ Age + Disease_status, data = combined_df%>%filter(Disease_status=="Healthy"|ID==test_ID),random = ~ 1|ID)
  return(lmer.indidividual)
})
all_lmer_summaries.indel<-lapply(all_individual_lmer_list.indel,summary)
summary_df.indel<-dplyr::bind_rows(lapply(all_lmer_summaries.indel,function(list) list$tTable["Disease_statusSCD",]))%>%
  mutate(CI_lower=Value-1.96*Std.Error,CI_upper=Value+1.96*Std.Error)%>%
  mutate(ID=paste0("SCD",1:6),.before=1)
print(summary_df.indel)

```

Model including age + an age:disease interaction term only

```{r}
lmer.age_and_disease_int.indel <- lmer(INDEL_burden_adj ~ Age + Age:Disease_status + (1|ID), data = combined_df)
summary(lmer.age_and_disease_int.indel)

```

Model including age + disease + an age:disease interaction term

```{r}
lmer.age_and_disease_plus_int.indel <- lmer(INDEL_burden_adj ~ Age*Disease_status + (1|ID), data = combined_df)
summary(lmer.age_and_disease_plus_int.indel)
```
